Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-31 doc/network.tex
updating directory
ugh more minor tweaks

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproduction number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

@@

During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t i(s) b_s(\tau) \,\mathrm{d}s.
\end{equation}
The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproduction number):
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} \mathcal R_{\rm c}(s) i(s) f_s(\tau) \,\mathrm{d}s.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The mean backward and aggregated generation interval are calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all infectious contacts by a focal individual, regardless of whether the contact results in infection.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions have a shorter mean than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution.
Since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the aggregated generation-interval distributions (based on the first 1000 realized generation intervals) without correcting for right-truncation (labeled as `contact tracing') severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 75 infected individuals, as well as the estimates based on the initial forward generation intervals, which are calculated by applying the Euler-Lotka equation to the realized generation intervals of all infections caused by the first 75 infected individuals.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

>>>>

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproductive number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

@@

During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t i(s) b_s(\tau) \,\mathrm{d}s.
\end{equation}
The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} \mathcal R_{\rm c}(s) i(s) f_s(\tau) \,\mathrm{d}s.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The mean backward and aggregated generation interval are calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all infectious contacts by a focal individual, regardless of whether the contact results in infection.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions have a shorter mean than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproductive number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution.
Since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the aggregated generation-interval distributions (based on the first 1000 realized generation intervals) without correcting for right-truncation (labeled as `contact tracing') severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 75 infected individuals, as well as the estimates based on the initial forward generation intervals, which are calculated by applying the Euler-Lotka equation to the realized generation intervals of all infections caused by the first 75 infected individuals.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}
